TacticToe: Learning to Prove with Tactics

### 2018-02-20 tactictoe_jv.tex
moving abstract to introduction

\todoi{Move part of the abstract to the introduction.}
\begin{abstract}
Formalization of mathematics in interactive theorem provers (ITPs) relies today on many proof automation methods.
Some of these methods are general-purpose and with enough time they can prove any provable conjecture.
Many formalized proof scripts however also take advantage of specialized rules and theory-based strategies.
In the HOL4 ITP system, practically all of them are implemented as \emph{tactics}.

Automated theorem provers (ATPs) rely on parameterizable general
strategies. Sometimes, a strategy can be specialized for only one theory as in
SMT solving. But it is difficult for an ATP to rely on rules tailored for a particular case.
Indeed, this would require implementing many specific rules and more
importantly figuring out in which cases they can be applied.

We bridge this gap by implementing our tactical prover TacticToe on top
of the HOL4 ITP.
TacticToe first learns from many proof scripts
the relevance of tactics in particular proof situations.
This learned knowledge is then used in a Monte Carlo tree search algorithm that explores the most promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\%
of the 7164 theorems in HOL4's standard library whereas
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the
results of both provers.
\end{abstract}

\tableofcontents
\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Such tactics typically analyze the current proof state (goal and
  assumptions), then apply non-trivial proof transformations that get
  expanded into many basic kernel-level inferences or large parts of a proof term.
%\marginpar{not Coq}
  In this work we develop a tactic-level automation procedure for the \holfour ITP
  which guides the selection of tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer
  systems~\cite{hammers4qed,tgck-cpp15}, our technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  searches directly for sequences of tactic applications that lead to
  an ITP proof, thus avoiding the translation and proof-reconstruction phases needed by the hammers.

  To do this, we \emph{extract and record} tactic invocations from the ITP
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as
  guidance for a \emph{Monte Carlo tree search} that explores different
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics.  The system is evaluated on a large set of theorems originating
  from \holfour (Section~\ref{s:experiments}).

@@

\paragraph{Goals and theorems.}
A goal is represented as a \emph{sequent}.
A sequent is composed of a set of assumptions and a conclusion, all of which are higher-order logic formulas.
When a goal is proven, it becomes a theorem.
If the developer gives the theorem a name then we refer to it as a top-level theorem.

@@

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search~\cite{montecarlo} (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and prettification of the generated proof improves user
interaction.
\end{itemize}

>>>>

\todoi{Move part of the abstract to the introduction.}
\begin{abstract}

We implement a tactical prover TacticToe on top of the HOL4 interactive theorem 
prover. TacticToe learns from many human proofs which mathematical method is 
useful in particular proof situations.
This knowledge is then used in a Monte Carlo tree search algorithm that 
explores the most promising tactic-level proof paths.
On a single CPU, with a time limit of 60 seconds, TacticToe proves 66.4\% of 
7164 theorems in HOL4's standard library whereas
Eprover solves 34.5\%. The success rate rises to 69.0\% by combining the
results of \tactictoe and \eprover.
\end{abstract}

\tableofcontents

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Such tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation: 
  specialized rules, theory-based strategies and general purpose startegies. 
  Thanks to the progress in proof 
  automation, developers can delegate more and more complicated proof 
  obligations to general purpose strategies. Those are implemented by automated 
  theorem provers(ATPs) such as \eprover~\cite{eprover}. The communications 
  between 
  an ITP and ATPs is made possible by a ``hammer''  
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing 
  premise selection, translation and proof reconstruction.
  Yet, ATPs are far from foolproof and more precise user-guidance, achieved 
  by applying a particular sequence of specialized rules, is almost always 
  necessary to develop a mathematical theory.

  In this work, we develop in the ITP \holfour a procedure to select suitable 
  tactics depending on the current proof state by learning 
  from previous proofs. Backed by this machine-learned guidance, our prover 
  \tactictoe relies on Monte 
  Carlo tree search~\cite{montecarlo} to find sequences of tactic applications 
  leading to an ITP proof.
  To this end, we \emph{extract and record} tactic invocations from the ITP
  proofs (Section~\ref{sec:recording}) and
  \emph{build efficient machine learning classifiers} based on such training
  examples (Section~\ref{s:prediction}).  The learned data serves as
  guidance for a \emph{Monte Carlo tree search} that explores different
  proof paths (Section~\ref{sec:proofsearch}). The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics (Section~\ref{sec:proofdisplay}). The system is evaluated on a large 
  set of theorems originating
  from \holfour (Section~\ref{s:experiments}).

@@

\paragraph{Goals and theorems.}
A goal (or proof state) is represented as a \emph{sequent}.
A sequent is composed of a set of assumptions and a conclusion, all of which are higher-order logic formulas.
When a goal is proven, it becomes a theorem.
If the developer gives the theorem a name then we refer to it as a top-level theorem.

@@

\begin{itemize}
\item Proof recording at the tactic level is made more precise. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm. The MCTS algorithm gives a more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner.
\item A tactic abstraction mechanism is introduced. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover to help \tactictoe during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library.
\item First evaluation of \tactictoe success rate relative to
the length of the original human proofs.
\item Minimization and prettification of the generated proof  facilitates user
interaction.
\end{itemize}
